A Constructive Ordered Predicate Logic \\ for Category Theory

### 2022-10-14 paper/paper.tex
spell check main paper

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex and hard to learn, especially for
computer scientists who do not have a strong background in pure
mathemantics.
%
While in some ways this could be remedied by better introductory
materials, in some ways category theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
For example, when constructing an adjoint pair of
functors between two categories, a na\"ive approach would define all of
the data of the action on objects, action on arrows, prove the
functoriality of such actions, as well as construct two families of
transformations, prove they are natural and then finally proving a pair
of equalities relating compositions of natural transformations.
%
Carrying out these proofs explicitly is quite tedious and many newcomers
to are left with the impression that category theory is full of long,
but ultimately trivial constructions.
%
This complexity is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
Moreover, these generalizations are often what is needed for programming
language applications; for example, domain- and metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

@@

The syntax of \vett{} is an indexed, ordered linear, proof-relevant
variant of predicate logic over a unary type theory.
%
Just as a predicate logic has a notion of type, term, relation and
implication, \vett{} is based on four analogous category theoretic
concepts: categories, functors, profunctors and natural transformations
of profunctors.
%
Categories are treated like types, and the unary functors we consider in
this paper are each represented by a term whose type is a category and
whose one free variable ranges over a category.
%
The analogue of a relation is a \emph{profunctor} (defined below), which
are written as a kind of set with free category variables.  Like the
restriction to unary functors, we restrict to profunctors with two free
variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
We can strenghten this analogy into a formal construction: any
construction in \vett{} can be erased to a corresponding construction
or proof in predicate logic, as sets, functions and relations define a 
(somewhat degenerate) virtual equipment.
%% %
%% Our syntax highlights this analogy in that some of our constructions
%% are generalizations of set comprehension or power-sets and we use
%% similar notation.

While the restricted syntax developed in this paper does not express
important concepts such as functor categories or opposite categories,
the restriction is natural in that it corresponds exactly to a
well-understood notion of model, virtual equipments, which can express a
great deal of fundamental results and constructions in category
theory~\cite{elements-of-oo-category-theory,
  enriched-indexed-categories}.
%
Moveover, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories, we can state a theorem that quantifies over functors using
the meta-language's ``external'' universal quantifier (which does not
have automatic functoriality/naturality properties).  To support this,
\vett includes a third layer, an extensional dependent type theory in
the style of Martin-L\"of type theory. All of our ordered predicate
logic judgments are also be indexed by a context from this dependent
type theory, and the type theory includes universe types for categories,
functors, profunctors and natural transformations.  This allow us to
formalize theorems the logic is too restrictive to encode, analogous to
2-level~\cite{voevodsky13hts,altenkirch+16strict} or indexed type
theories~\cite{isaev21indexed,cervesatopfenning02llf,vakar15linear,krishnaswami+15linear}.

@@

In Figure~\ref{fig:hol-cats} we give a table summarizing the
relationship between the judgments and connectives of higher-order
predicate logic with our ordered variant. Due to the incorporation of
variance, some unorderd concepts generalize to multiple different
ordered notions. For instance, covariant and contravariant presheaf
categories generalize the power set. Further, because we only have
binary relations rather than relations of arbitrary arity, we have
only restricted forms of universal and existential quantification
which come combined with implications and conjunctions.

@@

Next, we turn to some of the central theorems of category theory, the
Yoneda and Co-Yoneda lemmas. Despite being ulimately quite elementary,
these are notoriously abstract. In \ohol{}, we can view these as
ordered generalizations of some very simple tautologies about
equality. For instance, the Yoneda lemma generalizes
the equivalence between the formulae $\forall x. \forall y. x = y
\Rightarrow P y$ and $P x$.
\begin{lemma} \label{lem:yoneda}
  Let $\alpha^o : \cat C$ and $\pi : \negPresheaf {\cat C}$. Then
  \begin{enumerate}
  \item (Yoneda) The profunctor $\homrallXYtoZ {\alpha'} {\punitinXfromYtoZ {\cat C} {\alpha'}{\alpha}} {\negPresheafAppPtoX \pi {\alpha'}}$ is isomorphic to $\negPresheafAppPtoX \pi \alpha$
  \item (Co-Yoneda) The profunctor $\tensorexistsXwithYandZ {\alpha'}{\punitinXfromYtoZ{}{\alpha}{\alpha'}} {\negPresheafAppPtoX \pi \alpha}$ is isomorphic to $\negPresheafAppPtoX \pi \alpha$
  \end{enumerate}
\end{lemma}
The proofs both follow from the unit elimination rule, which is
essential the Yoneda lemma---the two cases of showing (1) is an
isomorphism are precisely the $\beta$ and $\eta$ rules for the unit.

@@

We now comapre \vett{} with other calculi for formal category theory.

Riehl and Verity~\cite{rvXXelements} use a formal language of virtual
equipments to prove results valid for $\infty$-categories without
concrete manipulation of model categories. They formalize this language
as a theory in Makkai's framework of first-order logic with dependent
sorts (FOLDS).  While this previous work has the same models as \vett{},
we believe that the syntax we propose in this paper formalizes informal
arguments more directly, as shown in Section~\ref{sec:examples}.  This
is because FOLDS approach approach is entirely relational, whereas we
formalize concepts like restriction of a profunctor or composition of
natural transformations as functional operations (substitution).  In
particular, this means that our calculus requires only vertically
degenerate squares (elements/natural transformations) as a
``user-facing'' notion, with general squares occuring only in the
admissible substitution operations.

@@

\textbf{Acknowledgements.}  This material is based on research sponsored
by the National Science Foundation under agreement number CCF-1909517
and the United States Air Force Research Laboratory under agreement
number FA9550-21-0009 (Tristan Nguyen, program manager).
%% The U.S. Government is authorized to reproduce and distribute
%% reprints for Governmental purposes notwithstanding any copyright
%% notation thereon.  The views and conclusions contained herein are
%% those of the authors and should not be interpreted as necessarily
%% representing the official policies or endorsements, either expressed
%% or implied, of the National Science Foundation, the United States Air
%% Force Research Laboratory, or the U.S. Government.
The authors would like to thank David Jaz Myers, Emily Riehl, Mike
Shulman, Dominic Verity for helpful feedback on this work.

>>>>

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex and hard to learn, especially for
computer scientists who do not have a strong background in pure
mathematics.
%
While in some ways this could be remedied by better introductory
materials, in some ways category theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
For example, when constructing an adjoint pair of
functors between two categories, a na\"ive approach would define all of
the data of the action on objects, action on arrows, prove the
functoriality of such actions, as well as construct two families of
transformations, prove they are natural and then finally proving a pair
of equalities relating compositions of natural transformations.
%
Carrying out these proofs explicitly is quite tedious and many newcomers
to are left with the impression that category theory is full of long,
but ultimately trivial constructions.
%
This complexity is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
Moreover, these generalizations are often what is needed for programming
language applications; for example, domain- and metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

@@

The syntax of \vett{} is an indexed, ordered linear, proof-relevant
variant of predicate logic over a unary type theory.
%
Just as a predicate logic has a notion of type, term, relation and
implication, \vett{} is based on four analogous category theoretic
concepts: categories, functors, profunctors and natural transformations
of profunctors.
%
Categories are treated like types, and the unary functors we consider in
this paper are each represented by a term whose type is a category and
whose one free variable ranges over a category.
%
The analogue of a relation is a \emph{profunctor} (defined below), which
are written as a kind of set with free category variables.  Like the
restriction to unary functors, we restrict to profunctors with two free
variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
We can strengthen this analogy into a formal construction: any
construction in \vett{} can be erased to a corresponding construction
or proof in predicate logic, as sets, functions and relations define a 
(somewhat degenerate) virtual equipment.
%% %
%% Our syntax highlights this analogy in that some of our constructions
%% are generalizations of set comprehension or power-sets and we use
%% similar notation.

While the restricted syntax developed in this paper does not express
important concepts such as functor categories or opposite categories,
the restriction is natural in that it corresponds exactly to a
well-understood notion of model, virtual equipments, which can express a
great deal of fundamental results and constructions in category
theory~\cite{elements-of-oo-category-theory,
  enriched-indexed-categories}.
%
Moreover, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories, we can state a theorem that quantifies over functors using
the meta-language's ``external'' universal quantifier (which does not
have automatic functoriality/naturality properties).  To support this,
\vett includes a third layer, an extensional dependent type theory in
the style of Martin-L\"of type theory. All of our ordered predicate
logic judgments are also be indexed by a context from this dependent
type theory, and the type theory includes universe types for categories,
functors, profunctors and natural transformations.  This allow us to
formalize theorems the logic is too restrictive to encode, analogous to
2-level~\cite{voevodsky13hts,altenkirch+16strict} or indexed type
theories~\cite{isaev21indexed,cervesatopfenning02llf,vakar15linear,krishnaswami+15linear}.

@@

In Figure~\ref{fig:hol-cats} we give a table summarizing the
relationship between the judgments and connectives of higher-order
predicate logic with our ordered variant. Due to the incorporation of
variance, some unordered concepts generalize to multiple different
ordered notions. For instance, covariant and contravariant presheaf
categories generalize the power set. Further, because we only have
binary relations rather than relations of arbitrary arity, we have
only restricted forms of universal and existential quantification
which come combined with implications and conjunctions.

@@

Next, we turn to some of the central theorems of category theory, the
Yoneda and Co-Yoneda lemmas. Despite being ultimately quite elementary,
these are notoriously abstract. In \ohol{}, we can view these as
ordered generalizations of some very simple tautologies about
equality. For instance, the Yoneda lemma generalizes
the equivalence between the formulae $\forall x. \forall y. x = y
\Rightarrow P y$ and $P x$.
\begin{lemma} \label{lem:yoneda}
  Let $\alpha^o : \cat C$ and $\pi : \negPresheaf {\cat C}$. Then
  \begin{enumerate}
  \item (Yoneda) The profunctor $\homrallXYtoZ {\alpha'} {\punitinXfromYtoZ {\cat C} {\alpha'}{\alpha}} {\negPresheafAppPtoX \pi {\alpha'}}$ is isomorphic to $\negPresheafAppPtoX \pi \alpha$
  \item (Co-Yoneda) The profunctor $\tensorexistsXwithYandZ {\alpha'}{\punitinXfromYtoZ{}{\alpha}{\alpha'}} {\negPresheafAppPtoX \pi \alpha}$ is isomorphic to $\negPresheafAppPtoX \pi \alpha$
  \end{enumerate}
\end{lemma}
The proofs both follow from the unit elimination rule, which is
essential the Yoneda lemma---the two cases of showing (1) is an
isomorphism are precisely the $\beta$ and $\eta$ rules for the unit.

@@

We now compare \vett{} with other calculi for formal category theory.

Riehl and Verity~\cite{rvXXelements} use a formal language of virtual
equipments to prove results valid for $\infty$-categories without
concrete manipulation of model categories. They formalize this language
as a theory in Makkai's framework of first-order logic with dependent
sorts (FOLDS).  While this previous work has the same models as \vett{},
we believe that the syntax we propose in this paper formalizes informal
arguments more directly, as shown in Section~\ref{sec:examples}.  This
is because FOLDS approach approach is entirely relational, whereas we
formalize concepts like restriction of a profunctor or composition of
natural transformations as functional operations (substitution).  In
particular, this means that our calculus requires only vertically
degenerate squares (elements/natural transformations) as a
``user-facing'' notion, with general squares occurring only in the
admissible substitution operations.

@@

\textbf{Acknowledgments.}  This material is based on research sponsored
by the National Science Foundation under agreement number CCF-1909517
and the United States Air Force Research Laboratory under agreement
number FA9550-21-0009 (Tristan Nguyen, program manager).
%% The U.S. Government is authorized to reproduce and distribute
%% reprints for Governmental purposes notwithstanding any copyright
%% notation thereon.  The views and conclusions contained herein are
%% those of the authors and should not be interpreted as necessarily
%% representing the official policies or endorsements, either expressed
%% or implied, of the National Science Foundation, the United States Air
%% Force Research Laboratory, or the U.S. Government.
The authors would like to thank David Jaz Myers, Emily Riehl, Mike
Shulman, Dominic Verity for helpful feedback on this work.
